Adaptive Video Streaming with PID-Based Quality Control and Optimized Bitstream Extraction

### 2015-04-27 tmm.tex
section title shoule be capitalized
the version I think OK

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
In order to give the user best quality of service under bandwidth fluctuation, video streaming systems should be able to adjust the video quality according to available bandwidth. In this paper we propose such an adaptive video streaming system which adopts the Scalable Video Coding (SVC) technology and integrates novel models and algorithms. First, we utilize the Proportional-Integral-Derivative (PID) method to control the video quality adjustment according to actual bandwidth condition. By considering bandwidth information of the past, the current and even the future, the PID-based quality control algorithm is able to determine a suitable bitrate for transmission and deliver highest video quality with least fluctuation. Second, we present an efficient scheme to extract an optimal sub-stream from the whole SVC video bitstream under given bitrate constraint. A linear error model is used to accurately estimate the distortion caused by discarding any combination of data packets, and a greedy-like algorithm is designed to assign each data packet a priority value for Rate-Distortion optimized extraction. Experimental results show that the PID-based quality control algorithm brings 8.6\% higher video quality with 24.8\% lower quality fluctuation comparing with the baseline algorithm, and the bitstream extraction method achieves a PSNR gain of up to 0.5dB with no computational complexity increment. Altogether, the system produces a higher and smoother PSNR curve representing the final video quality perceived by the user. The proposed system is also implemented in online video website www.7dlive.com and proves to be practically effective.
\end{abstract}

@@

One of the biggest challenge for video streaming is the variety of users' network environment. A user may use different types of network (Ethernet, Wi-Fi, Cellular, etc) and may also switch from one type to another (e.g., from Wi-Fi to cellular network) on going. Even in the same network, the user's bandwidth may change from time to time, influenced by the overall traffic. During such an environment, bandwidth fluctuation is inevitable. If the video is streamed at a constant bitrate, it would be difficult or impossible for the user to get best watching experience. To cope with this challenge, the video streaming system should be able to automatically adjust the video's bitrate or quality according to the bandwidth change. In other words, the video streaming system needs to be adaptive.

@@

In the SVC video streaming system, one scalable video stream with enhancement data packets is produced according to actual requirement and stored at the server for streaming (how to configure and produce the stream is not in the scope of this paper; users are referred to \cite{JSVM}). During the streaming process, there are two key problems that need to be addressed well. The first is quality control, and the second is bitstream extraction. Quality control means controlling the video quality (or the video bitrate, from the transmission point of view) to follow the change of the bandwidth. Once the bandwidth has changed, the quality control algorithm should determine a bitrate suitable for the current bandwidth. Given this bitrate, how to extract data from the whole SVC stream to form an optimal sub-stream becomes the second problem, i.e., bitstream extraction problem. Optimized bitstream extraction aims to select a subset of NALUs (or packets) which represents video with best quality, or minimal distortion, under the constrain of the given bitrate. It should be noted that, bitstream extraction is not only useful in adaptive video streaming, but also in other SVC application scenarios where part of data needs to be obtained or kept, e.g., the archiving of surveillance video.

@@

To implement our proposed adaptive streaming system requires the combination of the PID-based quality control algorithm and the optimized bitstream extraction method. The former determines the current bitrate to follow the bandwidth change, and the latter will ensure the sub-stream with the best possible quality is obtained and sent under that bitrate constrain. They work together to construct a system that delivers best quality of experience for the users. Choosing DSS as the implementation basis, we have integrated the proposed algorithms into it for the purpose of both research and application. In this section, we first present separate experiments to demonstrate the advantage of each proposed solution, and then give the result when they are combined together. Readers who are interested in the practical implementation can also visit the online video website http://www.7dlive.com for more information.

@@

\subsection{Result of the combined system}

After combining the two algorithms together, the proposed adaptive streaming system is able to deliver a higher and smoother video quality for the user. We compare the proposed system's performance with DSS that supports SVC but has its original quality control method and JSVM Basic bitstream extraction. The result for the same test stream as used in subsection \ref{subsec:exp-control} is present in Fig. \ref{fig:fluctuation-psnr}. It shows PSNR change perceived by the user when available bandwidth fluctuates around 800 kbps. It can be seen that, the proposed system has effectively avoided unnecessary quality drop (note the curve around time 50s and 350s) and been able to keep a relatively higher PSNR value (note the curve between time 450s and 600s).

@@

In this paper, we propose an adaptive video streaming system with PID-based quality control algorithm and optimized bitstream extraction. The PID-based quality control algorithm has a flexible mechanism to determine the suitable quality level so that it gives an accurate and swift response to the bandwidth change. Comparing with the original algorithm in DSS, this algorithm has reduced quality fluctuation significantly to improve audience satisfaction. The bitstream extraction method includes a linear error model to estimate the distortion caused by discarding any combination of packets and a greedy algorithm to assign priority to each data packet. The packets' priority values can then be stored in the bitstream and used for R-D optimized extraction. Experimental results show that the proposed method can achieve PSNR gain of up to 0.5dB compared to the JSVM QL extractor under the same bitrate constrain, also with the same computational complexity. The SVC streaming system integrated with these two solutions performs well in automatically adapting to the user's bandwidth fluctuation, and is able to give the user a better watching experience.

>>>>

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
To cope with the challenge brought by bandwidth fluctuation and provide the user with best quality of service, this paper proposes an adaptive video streaming system which adopts the Scalable Video Coding (SVC) technology and integrates novel models and algorithms. First, the Proportional-Integral-Derivative (PID) method is utilized to control the video quality adjustment according to actual bandwidth condition. By considering bandwidth information of the past, the current and even the future, the PID-based quality control algorithm determines a suitable bitrate for transmission. Then, under the constraint of that bitrate, an efficient scheme is used to extract an optimal sub-stream from the whole SVC video bitstream. In this scheme, a linear error model enables accurate estimation of the distortion caused by discarding any combination of data packets, and a greedy-like algorithm is designed to assign each data packet a priority value for Rate-Distortion optimized bitstream extraction. Experimental results show that by combining the above algorithms, PSNR average can be increased from 40.29 to 40.62, and PSNR variance can be reduced from 0.66 to 0.28, i.e., both higher video quality and lower quality fluctuation are achieved. The proposed system is also implemented in online video website www.7dlive.com and proves to be practically effective.
\end{abstract}

@@

One of the biggest challenge for video streaming is the variety of users' network environment. A user may use different types of network (Ethernet, Wi-Fi, Cellular, etc.) and may also switch from one type to another (e.g., from Wi-Fi to cellular network) on going. Even in the same network, the user's bandwidth may change from time to time, influenced by the overall traffic. During such an environment, bandwidth fluctuation is inevitable. If the video is streamed at a constant bitrate, it would be difficult or impossible for the user to get best watching experience. To cope with this challenge, the video streaming system should be able to automatically adjust the video's bitrate or quality according to the bandwidth change. In other words, the video streaming system needs to be adaptive.

@@

In the SVC video streaming system, one scalable video stream with enhancement data packets is produced according to actual requirement and stored at the server for streaming (how to configure and produce the stream is not in the scope of this paper; users are referred to \cite{JSVM}). During the streaming process, there are two key problems that need to be addressed well. The first is quality control, and the second is bitstream extraction. Quality control means controlling the video quality (or the video bitrate, from the transmission point of view) to follow the change of the bandwidth. Once the bandwidth has changed, the quality control algorithm should determine a bitrate suitable for the current bandwidth. Given this bitrate, how to extract data from the whole SVC stream to form an optimal sub-stream becomes the second problem, i.e., bitstream extraction problem. Optimized bitstream extraction aims to select a subset of NALUs (or packets) which represents video with best quality, or minimal distortion, under the constraint of the given bitrate. It should be noted that, bitstream extraction is not only useful in adaptive video streaming, but also in other SVC application scenarios where part of data needs to be obtained or kept, e.g., the archiving of surveillance video.

@@

To implement our proposed adaptive streaming system requires the combination of the PID-based quality control algorithm and the optimized bitstream extraction method. The former determines the current bitrate to follow the bandwidth change, and the latter will ensure the sub-stream with the best possible quality is obtained and sent under that bitrate constraint. They work together to construct a system that delivers best quality of experience for the users. Choosing DSS as the implementation basis, we have integrated the proposed algorithms into it for the purpose of both research and application. In this section, we first present separate experiments to demonstrate the advantage of each proposed solution, and then give the result when they are combined together. Readers who are interested in the practical implementation can also visit the online video website http://www.7dlive.com for more information.

@@

\subsection{Result of the Combined System}

After combining the two algorithms together, the proposed adaptive streaming system is able to deliver a higher and smoother video quality for the user. We compare the proposed system's performance with DSS that supports SVC but has its original quality control method and JSVM Basic bitstream extraction. The result for the same test stream as used in subsection \ref{subsec:exp-control} is present in Fig. \ref{fig:fluctuation-psnr}. It shows PSNR change perceived by the user when available bandwidth fluctuates around 800 kbps. It can be seen that, the proposed system has effectively avoided unnecessary quality drop (note the curve around time 50s and 350s) and been able to keep a relatively higher PSNR value (note the curve between time 450s and 600s). The average and variance of PSNR delivered by the proposed system are 40.62 and 0.28, respectively, while the values of the baseline system are 40.29 and 0.66.

@@

In this paper, we propose an adaptive video streaming system with PID-based quality control algorithm and optimized bitstream extraction. The PID-based quality control algorithm has a flexible mechanism to determine the suitable quality level so that it gives an accurate and swift response to the bandwidth change. Comparing with the original algorithm in DSS, this algorithm has reduced quality fluctuation significantly to improve audience satisfaction. The bitstream extraction method includes a linear error model to estimate the distortion caused by discarding any combination of packets and a greedy algorithm to assign priority to each data packet. The packets' priority values can then be stored in the bitstream and used for R-D optimized extraction. Experimental results show that the proposed method can achieve PSNR gain of up to 0.5dB compared to the JSVM QL extractor under the same bitrate constraint, also with the same computational complexity. The SVC streaming system integrated with these two solutions performs well in automatically adapting to the user's bandwidth fluctuation, and is able to give the user a better watching experience.
